---
output: github_document
---
```{r setup, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = '#>')
```

# pense R package

<!-- begin badges -->
[![CRAN\_Status\_Badge](https://www.r-pkg.org/badges/version/pense)](https://CRAN.R-project.org/package=pense)
```{r, results='asis', eval=TRUE}
branch_name <- tryCatch(system('git rev-parse --abbrev-ref HEAD', intern = TRUE),
                        error = function (...) { return('master') },
                        warning = function (...) { return('master') })
                        
cat(sprintf('[![Build Status](https://travis-ci.com/dakep/pense-rpkg.svg?branch=master)](https://travis-ci.com/dakep/pense-rpkg)', branch_name))
```
<!-- end badges -->

This R package implements the penalized elastic net S-estimator (PENSE) and the penalized M-step (PENSEM) as proposed in [Cohen Freue, et al. (2019)](https://projecteuclid.org/euclid.aoas/1574910036) as well as the adaptive extensions developed in [Kepplinger (2020)](https://hdl.handle.net/2429/75637).

## Usage

The main function for most users is `adapense_cv()`, which computes an adaptive PENSE fit and estimates prediction performance for many values of the penalization level.

### Example

```r
library(pense)

# Generate dummy data with n=50 observations and p=25 possible predictors
# (of which only the first 3 are truly relevant).
n <- 50
p <- 25
set.seed(123)
x <- matrix(rt(n * p, df = 5), ncol = p)
y <- x[, 1] + 0.5 * x[, 2] + 2 * x[, 3] + rt(n, df = 2)

# Compute an adaptive PENSE fit with hyper-parameters alpha=0.8, exponent=2,
# and 50 different values for the penalization level.
# Prediction performance of the 50 fits for different penalization levels
# is estimated with 5-fold cross-validation,
# repeated 10 times (the more the better!).
# On selected platforms use 2 CPU cores.
set.seed(123) # Setting the seed is suggested for reproducibility of the CV results.
fit <- adapense_cv(x, y, alpha = 0.9, cv_k = 5, cv_repl = 10, ncores = 2)

# Visualize the estimated prediction performance using plot()
plot(fit)

# Summarize the model with best prediction performance using summary()
summary(fit)

# Summarize the model with "almost as-good prediction performance" as the best
# model using summary(lambda = "se")
summary(fit, lambda = "se")
```

The user can also compute a non-adaptive PENSE fit by using `pense_cv()` in place of `adapense_cv()` or a PENSEM fit by using `pensem_cv()`.

### Using a parallel cluster
The cross-validation functions in the pense package (those ending in `_cv()`) also accept a parallel cluster instead of the `ncores=` argument.
Setting the seed ensures the results are reproducible and the same as if not using a cluster!

Continuing the example above, this could look like

```r
library(parallel)

# Set up the cluster using 3 CPUs on the local machine:
par_clust <- makeCluster(3)

set.seed(123) # Setting the seed is suggested for reproducibility of the CV results.
fit_with_cluster <- adapense_cv(x, y, alpha = 0.9, cv_k = 5, cv_repl = 10,
                                cl = par_clust)

stopCluster(par_clust)
```

Note that the `cl=` argument cannot be mixed with the `ncores=` argument.
Only one of them can be specified!

## Installation
To install the latest release from CRAN, run the following R code in the R console:
```r
install.packages("pense")
```

The most recent stable version as well as the developing version might not yet be available on CRAN.
These can be directly installed from github using the
[devtools](https://cran.r-project.org/package=devtools) package:
```r
# Install the most recent stable version:
devtools::install_github("dakep/pense-rpkg")
# Install the (unstable) develop version:
devtools::install_github("dakep/pense-rpkg", ref = "develop")
```

## References
- Kepplinger, D. (2020). *Robust estimation and variable selection in high-dimensional linear regression models.* University of British Columbia. [available online](https://hdl.handle.net/2429/75637).
- Cohen Freue, GV. Kepplinger D, Salibián-Barrera M, Smucler E. (2019). Robust elastic net estimators for variable selection and identification of proteomic biomarkers. _Annals of Applied Statistics_. 13(4). [available online](https://projecteuclid.org/euclid.aoas/1574910036).
